Future Evolution of CSMA Protocols for the IEEE 802.11 Standard

### 2013-03-01 paper.tex
working on the figure

Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions. The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA. Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
This value is roughly equal to the expectation of the random backoff used in the same situation in CSMA/CA.
This particular choice guarantees fairness between new stations and legacy stations.

@@

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  \colorbox{red}{$r \leftarrow 0$ ; $s \leftarrow 0$} \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    $s \leftarrow 0$ \;
	 \eIf{success}{
      %\tcc{Random backoff.}
      $b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$\;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/CA}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    $s \leftarrow 0$ \;    
    \eIf{success}{
      %\tcc{Random backoff.}
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/ECA}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of 1 packet \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    %$s \leftarrow 0$ \;
    \eIf{success}{
      %\tcc{Random backoff.}
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis}
\label{alg:CSMA_ECA_hist}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      Attempt transmission of $2^s$ packets \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    %$s \leftarrow 0$ \;
    \eIf{success}{
      %\tcc{Random backoff.}
      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis and fair-share}
\label{alg:CSMA_ECA_hist_fair}
\end{algorithm}

@@

To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced. It achieves greater throughput than CSMA/ECA with hysteresis only, while being fair for any number of contenders. This is achieved by using packet aggregation, transmitting $2^s$ packets per attempt. As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput. {Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network in terms of channel use.}

@@

Regarding the implementation in real prototypes, it has been proven difficult until now, as the protocols are embedded in the networks interfaces and it is not possible to reprogram them.
There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.

Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
RFID uses also a contention protocol which are prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.

>>>>

Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions. The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA. Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
This value is roughly equal to the expectation of the backoff chosen in line $18$ in Algorithm \ref{alg:CSMA_CA}.
This particular choice guarantees fairness between new stations and legacy stations.

@@

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  %\tcc{Initialize retransmission attempts.}
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  %\tcc{Initialize backoff counter.}
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet } \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    \colorbox{yellow}{$s \leftarrow 0$ } \;
	 \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/CA}
\label{alg:CSMA_CA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet } \;
      \If{collision}
      {
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    \colorbox{yellow}{$s \leftarrow 0$} \;    
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\caption{CSMA/ECA}
\label{alg:CSMA_ECA}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of 1 packet } \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    %$s \leftarrow 0$ \;
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis}
\label{alg:CSMA_ECA_hist}
\end{algorithm}

\begin{algorithm}[ht!!!]
\While{the device is on}
{
  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
  \While{there is a packet to transmit}{
    %\tcc{Initialize $a$.}
    \Repeat{($r = R$) or (success)}{
      %\tcc{First, backoff.}
      \While{$b>0$}{
        wait 1 slot \;
        $b \leftarrow b-1$ \;
      }
      \colorbox{yellow}{Attempt transmission of $2^s$ packets } \;
      \If{collision}{
        %\tcc{Random backoff.}
        $r \leftarrow r+1$ \;
        $s \leftarrow \min (s+1,S)$ \;
        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
      }
    }
    $r \leftarrow 0$ \;
    %$s \leftarrow 0$ \;
    \eIf{success}{
      %\tcc{Random backoff.}
      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
    }
    {
      Discard packet\;
      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
    }
  }
  Wait until there is a packet to transmit \;
}
\vspace{0.2cm}
\caption{CSMA/ECA with hysteresis and fair-share}
\label{alg:CSMA_ECA_hist_fair}
\end{algorithm}

An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff}.

\begin{figure*}[!t]
\centering
\includegraphics[width=7.0in]{figures/csma_eca_different_backoff}
\caption{Schedule length distributed adaption example. All the stations transmit the same number of packets in each cycle, despite using different schedule lengths.}
\label{fig:csma_eca_different_backoff}
\end{figure*}

@@

To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced. It achieves greater throughput than CSMA/ECA with hysteresis only, while being fair for any number of contenders. This is achieved by using packet aggregation, transmitting $2^s$ packets per attempt. As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput. {Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network.}

@@

Regarding the implementation in real prototypes, it has been proven difficult until now, as the IEEE 802.11 protocols are embedded in the network interfaces and it is not possible to reprogram them.
There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.

Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.
